
---
title: "A Myb-dominated gene regulatory network universally controls sexual cell fate transitions in diatoms"
author: "Gust Bilcke"
abstract: "Single-cell transcriptomic analyses that recreate key results and figures from: A Myb-dominated gene regulatory network universally controls sexual cell fate transitions in diatoms "
output:
  html_document:
    toc: true
    toc_depth: 4
    
---


# Continuous light (LL) experiments

## Download Seurat object from NCBI GEO

```{bash; eval = FALSE}

cd data/
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE303nnn/GSE303315/suppl/GSE303315%5FseuratObj%5Ffor%5Fpublication%2Erds

```

## Load Seurat object & get statistics

```{r}

# Required packages
library(slingshot); library(RColorBrewer); library(viridis); library(UCell); library(Seurat); library(openxlsx); library(ggplot2); library(paletteer); library(tradeSeq); library(dplyr); library(tidyr); library(ggnewscale); library(cowplot);  library(ggridges); library(clusterProfiler); library(ggbump)

# Seurat object (continuous light)
seuratObj <- readRDS("../data/GSE303315_seuratObj_for_publication.rds" )

# Number of cells
table(seuratObj@meta.data$Cluster)
table(seuratObj$Sample)
length(seuratObj$Sample)

# Number of UMIs per cell
mean(seuratObj$nCount_RNA) 
mean(seuratObj$nCount_RNA[seuratObj$Sample == "T12"]) 
mean(seuratObj$nCount_RNA[seuratObj$Sample == "T19"]) 
mean(seuratObj$nCount_RNA[seuratObj$Sample == "T24"]) 

# Total number of UMIs
sum(seuratObj$nCount_RNA) 
sum(seuratObj$nCount_RNA[seuratObj$Sample == "T12"]) 
sum(seuratObj$nCount_RNA[seuratObj$Sample == "T19"]) 
sum(seuratObj$nCount_RNA[seuratObj$Sample == "T24"]) 

# total Number of genes expressed (per set of cells)
RNA_data <- as.data.frame(seuratObj@assays$RNA@data)
expressed <- rowSums(RNA_data)
expressed_T12 <- rowSums(RNA_data[seuratObj$Sample == "T12"])
expressed_T19 <- rowSums(RNA_data[seuratObj$Sample == "T19"])
expressed_T24 <- rowSums(RNA_data[seuratObj$Sample == "T24"])
table(expressed > 0)
table(expressed_T12 > 0)
table(expressed_T19 > 0)
table(expressed_T24 > 0)

# average number of genes expressed / cell 
expr_vector <- colSums(RNA_data > 0)
mean(expr_vector)
mean(expr_vector[seuratObj$Sample == "T12"])
mean(expr_vector[seuratObj$Sample == "T19"])
mean(expr_vector[seuratObj$Sample == "T24"])

```

## Annotate UMAP

### Cell cycle phasing 

```{r}

CC <- read.xlsx("../data/cell_cycle_markers_Ccl.xlsx")
seuratObj <- CellCycleScoring(seuratObj, s.features = CC$gene_ID[CC$cell_cycle_stage == "G1 S" | CC$cell_cycle_stage == "S"], CC$gene_ID[CC$cell_cycle_stage == "M"],set.ident = F)

CCcols <- c("orange", "#E33D6F", "#297FD5", "tan")
CC_plot <- DimPlot(seuratObj, reduction="SCT_umap", group.by = "Phase", pt.size=0.22, cols = CCcols) + theme(legend.position = "none")  + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2")
CC_plot

```

### Annotate using bulk RNA-seq time series from Audoor et al. (https://onlinelibrary.wiley.com/doi/10.1111/mec.17320)

Plotting the average log2 fold changes of the top-20 top markers for each cluster

```{r}

# Load data
results <- read.xlsx("../data/Audoor_results_DE_Ccl.xlsx",1)[c(1,2,3,4)]
results <- results[results$contrast != "overall",]
allmarkers20 <- read.delim("../data/allMarkers20.txt")

# Analyze & plot
clusterlist <- list()
pal <- paletteer_c("ggthemes::Classic Red-Blue", 30)

for (cluster in allmarkers20$cluster){
  allmarkers20$gene2[allmarkers20$cluster == cluster] <- unlist(lapply(allmarkers20$gene[allmarkers20$cluster == cluster], function(x) gsub(x, pattern = "-", replacement = "_")))
  res_sel <- results[results$gene_ID %in% allmarkers20$gene2[allmarkers20$cluster == cluster],]
  res_agg <- aggregate(res_sel$LFC, by = list(res_sel$contrast), "mean")
  res_agg$cluster <- cluster
  clusterlist[[cluster]] <- res_agg
}

clusterexpr <- do.call(rbind, clusterlist)
colnames(clusterexpr) <- c("contrast", "LFC", "cluster")

clusterexpr$cluster <- factor(clusterexpr$cluster, levels = 1:16)
clusterexpr$contrast <- factor(clusterexpr$contrast, levels = c("T3", "T2", "T1"))

ggplot(clusterexpr) +
    geom_tile(aes(x = contrast, y = cluster, fill = LFC)) +
     scale_fill_gradientn(colours = rev(pal), limits = c(-max(abs(clusterexpr$LFC)),max(abs(clusterexpr$LFC))), na.value = "grey40") +
    theme_light() +
    theme(panel.grid = element_blank(), axis.ticks = element_blank()) +
    xlab("Bulk DE time point") +
    ylab("Single-cell cluster") +
  coord_flip()

```

### Annotate SIP (sex-inducing pheromone) response, using bulk RNA-seq data from Belisova, Bilcke et al. 2025 https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.20334 

```{r}


# Load data
SIP_9h <- read.delim("../data/SIP_9h_DE_Belisova2025.txt")
SIP_9h$gene <- unlist(lapply(SIP_9h$gene_ID, function(x) gsub("_", "-", x)))
sip_9hgenes <- SIP_9h$gene[SIP_9h$adjP < 0.05 & SIP_9h$logFC >0]

markers <- read.xlsx("../data/allmarkers_annotated.xlsx")

# Link SIP response to UMAP using UCELL
signatures4 <- list(sip_9hgenes = sip_9hgenes)
seuratObj <- AddModuleScore_UCell(seuratObj, features=signatures4, name = NULL)
FeaturePlot(seuratObj, reduction="SCT_umap", features = names(signatures4), pt.size=0.22)  + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

```

### Visualize different annotations on UMAP

```{r}

DimPlot(seuratObj, reduction="SCT_umap", group.by = "Sample", pt.size=0.22)+ labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

DimPlot(seuratObj, reduction="SCT_umap", group.by = "Cluster", pt.size=0.22)+ labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

DimPlot(seuratObj, reduction="SCT_umap", group.by = "Cell_type", pt.size=0.22)+ labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

DimPlot(seuratObj, reduction="SCT_umap", group.by = "Genotype", pt.size=0.22)+ labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

DimPlot(seuratObj, reduction="SCT_umap", group.by = "CellCyclePhase", pt.size=0.22)+ labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

```
## Trajectory inference

### Fit trajectory with Slingshot

```{r}

coordinates <- data.frame(seuratObj@reductions$SCT_umap@cell.embeddings[! seuratObj@meta.data$Cluster %in% c(3,2,11,8, 12,13),])
clusters <- seuratObj@meta.data$Cluster[! seuratObj@meta.data$Cluster %in% c(3,2,11, 8,12,13)]

sling <- slingshot(coordinates, clusters,start.clus = '1', end.clus = '16')

coordinates3 <- data.frame(seuratObj@reductions$SCT_umap@cell.embeddings)
pseudotimes <-slingPseudotime( slingshot::predict(sling, coordinates3))
weights <- slingCurveWeights( slingshot::predict(sling, coordinates3))
sling2 <-  slingshot::predict(sling, coordinates3)
colors <- colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)
plotcol <- colors[cut(pseudotimes, breaks=100)]

# pdf(file = "plots/Trajectory_inference/entire_trajectory_pseudotime2.pdf")
par(mar = c(1,1,1,1))
plot(coordinates3, col = plotcol,  pch=16, asp = 0.8, cex = 1)
lines(SlingshotDataSet(sling), lwd=5, col = 'black')
# dev.off()

# Save to Seurat Object
seuratObj@meta.data$pseudotime <- pseudotimes
seuratObj@meta.data$weights <- weights
FeaturePlot(seuratObj, reduction="SCT_umap",  pt.size=0.05, features = "pseudotime") +xlab("sctUMAP 1")+ylab("sctUMAP 2") +scale_color_gradientn(colours = viridis(n = 10)) + theme(legend.position = "none")

# Cell type composition along pseudotime
celltypecols <- c("#E64B35",  "tan4", "orange", "#90ba4c", "#477bd1")
ggplot() +
  geom_density(aes(x = seuratObj@meta.data$pseudotime[seuratObj@meta.data$Cell_type != "Stressed cells"], fill = seuratObj@meta.data$Cell_type[seuratObj@meta.data$Cell_type != "Stressed cells"]), position="fill") +
  scale_fill_manual(values = celltypecols) +
  theme_bw() +
    theme(panel.grid = element_blank()) +
    scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  xlab("pseudotime") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks.y = element_blank(), legend.title=element_blank())

```
#### Evaluate number of knots 

```{r; eval = FALSE}

counts <- as.matrix(seuratObj@assays$RNA@counts)

icMat <- evaluateK(counts =  counts,
          pseudotime = seuratObj@meta.data$pseudotime,
          cellWeights = seuratObj@meta.data$weights,
          nGenes = 100,
          k = 3:10  )

```

Relative AIC levels off at N=8 --> choose 8 knots

#### Fit generalized additive models (GAMs)

Run shell script at /scripts/associationTest/run_figGAM_cluster.sh

The output is saved in data/save_GAM_8knots.rdata

#### Perform associationtest (TradeSeq)

```{r}

load("../data/save_GAM_8knots.rdata") # Fitted GAM smoothers
assoc_test <- associationTest(models = gam, contrastType = "end", l2fc =sqrt(3), nPoints = 14) # npoints = 14 to avoid NAs

# Filtering
# Keep only those 8139 genes that are expressed in at least 50 cells with at least 2 UMIs. Loaded from a separate file for memory control
keepgenes <- as.character(read.table("../data/keep_genes_filtering.txt", header = T)$x)
assoc_test2 <- assoc_test[rownames(assoc_test) %in% keepgenes,] # 8139 kept

assoc_test2$adjP <- p.adjust(assoc_test2$pvalue, "BH")
sum(assoc_test2$adjP < 0.05 & ! is.na(assoc_test2$adjP)) # 3703 associated genes

# save(assoc_test2, file =  "../data/association_test2.RData")

```

## Mating type-specific expression 

### Test for genes with mating-type-specific expression

```{r}

seuratMT <- subset(seuratObj, subset = (Consensus_matingtype == "MT+" | Consensus_matingtype == "MT-"))
Idents(seuratMT) <- seuratMT@meta.data$Consensus_matingtype
allmarkers_MT <- FindAllMarkers(object = seuratMT, only.pos = TRUE, logfc.threshold = 3) 
table(allmarkers_MT$cluster[allmarkers_MT$p_val_adj < 0.05])
sig_MTmarkers <- allmarkers_MT[allmarkers_MT$p_val_adj < 0.05,]

```

### Clustering expression of mating types over pseudotime

```{r}

load("../data/predicted_MT_GAMs.RData") # Predicted expression on 60 pseudotime points per mating type   
sig_MTmarkers <- read.xlsx("../data/MT_genes_annotation.xlsx", 1)

predicted_plus <- df[df$lineage == "MT+" & df$gene_ID %in% sig_MTmarkers$gene[sig_MTmarkers$cluster == "MT+"],] # Cluster only on the pattern of the "dominant" MT
predicted_min <- df[df$lineage == "MT-" & df$gene_ID %in% sig_MTmarkers$gene[sig_MTmarkers$cluster == "MT-"],] # Cluster only on the pattern of the "dominant" MT

# Clustering the plus genes
k = 4
plus_wide <- spread(predicted_plus[c(1,3,4)], key = "gene_ID", value = "expr")
rownames(plus_wide) <- plus_wide$pseudotime
plus_wide <- plus_wide[-1]
dist_plus <- as.dist(1-cor(plus_wide))
clust_plus <- hclust(dist_plus)
clusters_plus <- cutree(clust_plus, k = k)
plot(clust_plus)

# Clustering the minus genes
min_wide <- spread(predicted_min[c(1,3,4)], key = "gene_ID", value = "expr")
rownames(min_wide) <- min_wide$pseudotime
min_wide <- min_wide[-1]
dist_min <- as.dist(1-cor(min_wide))
clust_min <- hclust(dist_min)
clusters_min <- cutree(clust_min, k = k)
plot(clust_min)

```

### Define function to plot genes over mating type-specific pseudotime

```{r}
plotMT_GAM_small <- function(gene_IDs){
  df2 <- df[df$gene_ID %in% gene_IDs,]
  ggplot(df2) +
  geom_line(aes(x = pseudotime, y = expr, colour = lineage, group = interaction(lineage,gene_ID)), size = 0.2) +
      scale_colour_manual(values = c("#439fb7", "#eb6349")) +
    new_scale_colour() +
  stat_summary(aes(x = pseudotime, y = expr, colour = lineage), geom = "line", fun = "mean", size = 3) +
          scale_colour_manual(values = c("#368195", "#d03517")) +
  ylab("Relative expression") +
  xlab("Pseudotime") +
  theme_bw() +
    theme(panel.grid = element_blank(), panel.background = element_rect(fill = "grey90")) +
 scale_y_continuous(limits = c(0, NA)) +
    scale_y_continuous(expand = expansion( mult = c(0, .05)))
}
```

### Extract and plot clusters from the hierarchical clustering

```{r}
plotlist <- list()

clusters_plus <- cutree(clust_plus, k = k)
clusters_plus[clusters_plus == 1] <- "3Midlate"
clusters_plus[clusters_plus == 2] <- "1Early"
clusters_plus[clusters_plus == 3] <- "2Middle"
clusters_plus[clusters_plus == 4] <- "4Late"

for (i in c("1Early", "2Middle", "3Midlate", "4Late")) {
  plot(plotMT_GAM_small(names(clusters_plus)[clusters_plus == i]) + theme(legend.position = "none"))
}

clusters_min <- cutree(clust_min, k = k)
clusters_min[clusters_min == 1] <- "4Late"
clusters_min[clusters_min == 2] <- "1Early"
clusters_min[clusters_min == 3] <- "2Middle"
clusters_min[clusters_min == 4] <- "3Midlate"

for (i in c("1Early", "2Middle", "3Midlate", "4Late")) {
  plot(plotMT_GAM_small(names(clusters_min)[clusters_min == i]) + theme(legend.position = "none"))
}
```

## Gene regulation

### Identify transcription factors (TFs) associated with pseudotime

```{r}

# First get association test data for all putative TFs (i.e. those with DNA binding domains)
putativeTFs <- read.delim("../data/C_closterium_TFs_added.txt")
load("../data/association_test2.RData")
associatedgenes <- rownames(assoc_test2[assoc_test2$adjP < 1e-5 & !is.na(assoc_test2$adjP),])
associatedTFs <- putativeTFs[putativeTFs$gene %in% associatedgenes,] # 58 associated putative TFs

# Filter TFs: Because this figure is a crucial starting point to pinpoint the TFs involved in sex, putative TFs were manually curated
checkTFs <- read.xlsx("../data/manual_curation_associatedTFs.xlsx") 
associatedTFs_checked <- associatedTFs[associatedTFs$gene %in% checkTFs$gene[checkTFs$conclusion == "TF"],] # 33 bona fide associated TFs
associatedTFs_checked$TF <- unlist(lapply(associatedTFs_checked$gene, function(x) checkTFs$newname[checkTFs$gene == x]))

```

### Plot TF ridgeline plots

```{r}

pal_pseudotimes <- paletteer_d("ggthemes::Tableau_20")[seq(1,19, by = 2)]
genepseudo <- read.delim("../data/gene_ordering_pseudotime.txt")$x # this file is sorted reversely (auxospores first, then ending with vegetative)!

counts <- as.matrix(seuratObj@assays$RNA@counts)
counts <- as.data.frame(counts)

plot_ridges <- function(df) {
  gene_IDs <- df$gene
  selectcounts <- counts[rownames(counts) %in% gene_IDs,]
  selectcounts$gene_ID  <- rownames(selectcounts)
  selectcounts$gene_ID <- factor(selectcounts$gene_ID, levels = genepseudo)
  selectcounts$family <- unlist(lapply(selectcounts$gene_ID, function(x) df$TF[df$gene == x]))
  selectL <- gather(selectcounts, 1:8674, key = "cell", value = "UMIs" )
  selectL$pseudotime <- rep(seuratObj@meta.data$pseudotime, each = length(gene_IDs) )

  for (gene_ID in unique(selectL$gene_ID)){
  x <- predict(loess(selectL$UMIs[selectL$gene_ID == gene_ID]~selectL$pseudotime[selectL$gene_ID == gene_ID], span = 0.3))
  selectL$smooth[selectL$gene_ID == gene_ID] <- x/max(x)
  }
  
  genepseudo2 <- unlist(lapply(genepseudo, function(x) gsub("-", "_", x)))
  selectL$gene <- unlist(lapply(selectL$gene_ID, function(x) gsub("-", "_", x)))
  selectL$gene <- factor(selectL$gene, levels = genepseudo2)

  ggplot(selectL)+
    geom_hline(yintercept = 0) +
    geom_ridgeline(aes(x = pseudotime, y = gene , height = smooth, fill = family), scale = 1.5) +
    ylab("TFs significantly associated with pseudotime") +
    xlab("Pseudotime") +
    facet_grid(family ~ ., space = "free", scales = "free") +
    scale_fill_manual(values = pal_pseudotimes) +
            theme_classic() +
        theme(strip.text.y = element_text(angle = 0, hjust = 0), strip.background = element_blank(), legend.position = "none", axis.line.y = element_blank(), axis.ticks.y = element_blank()) +
    scale_y_discrete(expand = c(0.05, 0)) +
  coord_cartesian(clip = "off")

  }

associatedTFs_checked$TF <- factor(associatedTFs_checked$TF, levels = c("bZIP", "TCX / CXC", "E2F family", "Myb", "Homeodomain", "HSF", "C2H2 ZnF", "CSF", "Aureochrome"))

plot_ridges(associatedTFs_checked)

```

### Gene regulatory network inference with MINI-EX

#### First export count table for Mini-ex

```{r; eval = FALSE}

write.table(counts, "data/MINI-EX/Ccl_count_matrix_scRNAseq_Ccl.txt", row.names = F, quote = F, sep = "\t")

```

#### Run MINI-EX

MINI-EX config file is stored in scripts/mini_ex_SLURM_LL_run.config
MINI-EX input data files are stored in data/MINI-EX

On a cluster system, run MINI-EX v1 using NextFlow like this:
```
nextflow -C scripts/mini_ex_SLURM_LL_run.config run miniex.nf 
```

### Explore gene regulatory network

#### Load GRN data

TF: transcription factor
TG: target gene

```{r}

ranked_regulons <- read.xlsx("../data/MINI-EX_output/Ccl_rankedRegulons.xlsx")
regulons_TGs <- read.delim("../data/MINI-EX_output/Ccl_regulons.txt", header = F)
regulons_TGs_long <- separate_rows(regulons_TGs, "V3", sep = ",")
colnames(regulons_TGs_long) <- c("TF", "cluster", "TG")
regulons_TGs_long$TF_cluster <- paste0(regulons_TGs_long$TF, "_", regulons_TGs_long$cluster)

length(unique(ranked_regulons$TF)) # 46 regulons driven by 27 TFs
length(unique(regulons_TGs_long$TG)) # 933 genes driven in those regulons

# Most regulons contain between 100 and 300 TGs
ggplot() +
  geom_histogram(aes(x = ranked_regulons$`#TGs`)) +
  theme_bw() +
  xlab("# TGs per regulon")

```

## Functional enrichment analysis of TGs within regulons

Using the enricher() function from ClusterProfiler (hypergeometric tests)

The background for hypergeometric tests contains homologous gene families (HOM), interpro domains (IPR), mating type specific genes, and cell cycle stages

```{r}

background <- read.delim("../data/full_enrichment_background.txt")

enrichlist <- list()
for (regulon in unique(regulons_TGs_long$TF_cluster)){
  enrichlist[[regulon]] <- cbind(as.data.frame(enricher(regulons_TGs_long$TG[regulons_TGs_long$TF_cluster == regulon], TERM2GENE = background, minGSSize = 1, maxGSSize = 19390, pAdjustMethod = "BH")@result), regulon) 
}

enrichdf2 <- do.call(rbind, enrichlist)
enrichdf3 <- enrichdf2[enrichdf2$Count > 3 & enrichdf2$qvalue < 0.05,]

```

Interpretation enrichment table ("enrichdf3"): M-phase (meiosis) is enriched among the target genes of Myb regulons in Clusters 6 and 7

# Diurnal cycle (LD) experiments

## Download Seurat object from NCBI GEO

```{bash; eval = FALSE}

cd data/
wget https://ftp.ncbi.nlm.nih.gov/geo/series/GSE303nnn/GSE303315/suppl/GSE303315%5FseuratObj2%5Ffor%5Fpublication%2Erds

```

## Get statistics

```{r}

# Seurat object (continuous light)
seurat_diurnal <- readRDS("../data/GSE303315_seuratObj2_for_publication.rds" )

# Number of cells
table(seurat_diurnal@meta.data$Experiment)

# Number of UMIs per cell 
mean(seurat_diurnal$nCount_RNA[seurat_diurnal$Experiment == "Light"]) # 4147.889
mean(seurat_diurnal$nCount_RNA[seurat_diurnal$Experiment == "Dark"]) # 3232.512

# Total number of UMIs 
sum(seurat_diurnal$nCount_RNA[seurat_diurnal$Experiment == "Light"]) # 22,070,920
sum(seurat_diurnal$nCount_RNA[seurat_diurnal$Experiment == "Dark"]) # 28,261,856

# total Number of genes expressed (per set of cells)
RNA_data <- as.data.frame(GetAssayData(seurat_diurnal, assay = "RNA", layer = "counts"))
expressed <- rowSums(RNA_data)
expressed_L <- rowSums(RNA_data[seurat_diurnal$Experiment == "Light"])
expressed_D <- rowSums(RNA_data[seurat_diurnal$Experiment == "Dark"])
table(expressed > 0) # 19,949 genes are expressed in total
table(expressed_L > 0) # 19,514 
table(expressed_D > 0) # 19,604 

# average number of genes expressed / cell
expr_vector <- colSums(RNA_data > 0)
mean(expr_vector) # 1701
mean(expr_vector[seurat_diurnal$Experiment == "Light"]) # 1984
mean(expr_vector[seurat_diurnal$Experiment == "Dark"]) # 1529

# cells per cell type
table(seurat_diurnal@meta.data$Experiment, seurat_diurnal@meta.data$Cell_type)

```

## Plot features on unprojected UMAP

```{r}

Idents(seurat_diurnal) <- seurat_diurnal@meta.data$Cluster

clustercols = c("#e6194b", "#3cb44b", "grey30", "orange1", "#4363d8", "violetred", "wheat3", "#f58231", "seagreen3", "violetred2", "gold1", "tomato2", "tan3", "cornflowerblue", "violet", "yellowgreen", "grey", "hotpink4")

celltypecols <- c("Vegetative / early signaling" = "#E64B35", "Stressed cells" = "grey70" ,"Mate finding" = "tan4", "Paired cells"="orange", "Gametes" = "#90ba4c","Zygotes & auxospores" = "#477bd1")

DimPlot(seurat_diurnal, reduction="SCT_umap", label=TRUE, label.size= 5, pt.size=0.22, cols = clustercols) + xlab("sctUMAP 1") + ylab("sctUMAP 2") 

DimPlot(seurat_diurnal, reduction="SCT_umap", group.by = "Experiment", pt.size=0.22, cols = c("steelblue4", "gold2")) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

DimPlot(seurat_diurnal, reduction="SCT_umap", group.by = "Cell_type", pt.size=0.22, cols = celltypecols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

```
## Plot features on projected UMAP (original UMAP of LL experiments)

```{r}

p <- DimPlot(seurat_diurnal, reduction="ref.umap", label=F, pt.size=0.3, cols = clustercols) + xlab("sctUMAP 1") + ylab("sctUMAP 2") 
LabelClusters(p, id = "ident",  repel = F,  label.size= 3)

DimPlot(seurat_diurnal, reduction="ref.umap", group.by = "Experiment", pt.size=0.18, cols = c("steelblue4", "gold2")) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

DimPlot(seurat_diurnal, reduction="ref.umap", group.by = "Cell_type", pt.size=0.3, cols = celltypecols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

```

## Formal analysis of top TFs for each library

Goal: Identify markers for gametogenesis for each of the 5 scRNA-seq libraries seperately: LL 12h, LL 19h, LL 24h, LD Light, LD Dark

### Define Gametogenesis cells in each dataset

```{r}

seuratObj@meta.data$Cell_type2 <- "Other"
seuratObj@meta.data$Cell_type2[seuratObj@meta.data$Cluster %in% c(6,7,8,9)] <- "Gametogenesis"
DimPlot(seuratObj, reduction="SCT_umap", group.by = "Cell_type2", pt.size=0.3) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

seurat_diurnal@meta.data$Cell_type2 <- "Other"
seurat_diurnal@meta.data$Cell_type2[seurat_diurnal@meta.data$Cluster == "J"] <- "Gametogenesis"
DimPlot(seurat_diurnal, reduction="ref.umap", group.by = "Cell_type2", pt.size=0.3) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

```


### subset both LL and DL data into seperate experiments

```{r}

# Create 5 independent Seurat objects
T12 <- subset(x = seuratObj, subset = Sample == "T12")
T19 <- subset(x = seuratObj, subset = Sample == "T19")
T24 <- subset(x = seuratObj, subset = Sample == "T24")
Light <- subset(x = seurat_diurnal, subset = Experiment == "Light")
Dark <- subset(x = seurat_diurnal, subset = Experiment == "Dark")

# Check with UMAPs that the subsetting worked
clustercols = c("#e6194b", "#3cb44b", "grey30", "orange1", "#4363d8", "violetred", "wheat3", "#f58231", "seagreen3", "violetred2", "gold1", "tomato2", "tan3", "cornflowerblue", "violet", "yellowgreen", "grey", "hotpink4")

DimPlot(T12, reduction="SCT_umap", pt.size=0.3, cols = clustercols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 
DimPlot(T19, reduction="SCT_umap", pt.size=0.3, cols = clustercols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 
DimPlot(T24, reduction="SCT_umap", pt.size=0.3, cols = clustercols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 
DimPlot(Light, reduction="ref.umap", pt.size=0.3, cols = clustercols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 
DimPlot(Dark, reduction="ref.umap", pt.size=0.3,  cols = clustercols) + labs(title = "")+xlab("sctUMAP 1")+ylab("sctUMAP 2") 

```

### Find markers for gametogenesis in each experiment & subset for TFs

```{r}

# Calculate markers for GAMETOGENESIS vs ALL OTHER CELLS
T12_markers_paired <- FindMarkers(T12, ident.1 = "Gametogenesis", group.by = "Cell_type2")
T19_markers_paired <- FindMarkers(T19, ident.1 = "Gametogenesis", group.by = "Cell_type2")
T24_markers_paired <- FindMarkers(T24, ident.1 = "Gametogenesis", group.by = "Cell_type2")
Light_markers_paired <- FindMarkers(Light, ident.1 = "Gametogenesis", group.by = "Cell_type2")
Dark_markers_paired <- FindMarkers(Dark, ident.1 = "Gametogenesis", group.by = "Cell_type2") #  despite having very few "paired cells" in the Dark sample, we can still find marker genes!

# Subset for only the TFs & clean the TF categories
TFs <- read.delim("../data/C_closterium_TFs_added.txt")
TFs <- TFs[! TFs$TF %in% c("SAP", "GARP", "LITAF (Zn)", "Sigma-70", "BolA", "Nin-like", "Tub ", "TAF", "MIZ (Zn)"),] # These are most likely not TF families
TFs$TF[TFs$TF == " Myb SHAQKYF"] <- "Myb"
TFs$TF[TFs$TF == "CCCH Zn finger"] <- "C3H ZnF"
TFs$TF[TFs$TF == "Homeodomain"] <- "HD/PHD"
TFs$TF[TFs$TF == "C2H2 Zn finger"] <- "C2H2 ZnF"
TFs$TF[TFs$TF == "CBF/NF-Y/archaeal"] <- "NF-Y"
TFs$TF[TFs$TF == "CPP (TCR)"] <- "CPP"
TFs$TF[TFs$TF == "E2F/DP"] <- "E2F"
TFs$TF[TFs$TF == "TATA-binding (Zn)"] <- "TATA-binding"

# Subset for TFs in Paired cells that are UPREGULATED
T12_TFs_paired <- T12_markers_paired[rownames(T12_markers_paired) %in% TFs$gene & T12_markers_paired$avg_log2FC > 0,]
T19_TFs_paired <- T19_markers_paired[rownames(T19_markers_paired) %in% TFs$gene & T19_markers_paired$avg_log2FC > 0,]
T24_TFs_paired <- T24_markers_paired[rownames(T24_markers_paired) %in% TFs$gene & T24_markers_paired$avg_log2FC > 0,]
Light_TFs_paired <- Light_markers_paired[rownames(Light_markers_paired) %in% TFs$gene & Light_markers_paired$avg_log2FC > 0,]
Dark_TFs_paired <- Dark_markers_paired[rownames(Dark_markers_paired) %in% TFs$gene & Dark_markers_paired$avg_log2FC > 0,]

T12_TFs_paired <- tibble::rownames_to_column(T12_TFs_paired, "gene")
T19_TFs_paired <- tibble::rownames_to_column(T19_TFs_paired, "gene")
T24_TFs_paired <- tibble::rownames_to_column(T24_TFs_paired, "gene")
Light_TFs_paired <- tibble::rownames_to_column(Light_TFs_paired, "gene")
Dark_TFs_paired <- tibble::rownames_to_column(Dark_TFs_paired, "gene")

library(dplyr)
T12_TFs_paired <- left_join(T12_TFs_paired, TFs, by = c("gene" = "gene"))
T19_TFs_paired <- left_join(T19_TFs_paired, TFs, by = c("gene" = "gene"))
T24_TFs_paired <- left_join(T24_TFs_paired, TFs, by = c("gene" = "gene"))
Light_TFs_paired <- left_join(Light_TFs_paired, TFs, by = c("gene" = "gene"))
Dark_TFs_paired <- left_join(Dark_TFs_paired, TFs, by = c("gene" = "gene"))

# Get statistics
Reduce(intersect, list(T12_TFs_paired$hits[1:9],T19_TFs_paired$hits[1:9], T24_TFs_paired$hits[1:9], Light_TFs_paired$hits[1:9], Dark_TFs_paired$hits[1:9]))

```

7/9 of the top-9 TFs are shared between every independent experiment

## Plot bump-plot of the top TFs for each experiment

```{r}

paired_tfs <- rbind(T12_TFs_paired[1:9,],T19_TFs_paired[1:9,],T24_TFs_paired[1:9,],Light_TFs_paired[1:9,],Dark_TFs_paired[1:9,])
paired_tfs$experiment <- c(rep("T12", 9),rep("T19", 9),rep("T24", 9),rep("Light", 9),rep("Dark", 9))
paired_tfs$rank <- rep(factor(1:9, levels = 9:1), 5)
paired_tfs$experiment <- factor(paired_tfs$experiment, levels = c("T12", "T19","T24", "Light", "Dark"))

paired_tfs$gene <- gsub("Ccl-", "Ccl_", paired_tfs$gene)

ggplot(paired_tfs) +
    geom_bump(aes(x = experiment, xend = experiment, y = rank, yend = rank, group = gene), size = 3, color = "grey", smooth = 9) +
  geom_label(aes(x = experiment, y = rank, fill = TF, label = gene), label.padding = unit(0.35, "lines"), size = 3.1) +
  theme_minimal() +
  scale_fill_manual(values = c(CSD = "#d37295", E2F = "#4bb42aff","HD/PHD" = "#22a4a2ff",HSF = "#e15759ff",Myb = "#b6992dff", "Fork head " = "yellow2", "C2H2 ZnF" = "#79706e", "C3H ZnF" = "white")) +
  theme(panel.grid = element_blank())

```



